GPU-Beschleunigung der Material Point Method

### 2019-01-27 mscthesis.tex
small fixed

\selectlanguage{english}
\vfill
\begin{center}
\subsubsection*{Abstract}
\end{center}
The Material-Point-Method is allowing for physically based simulations. It has found its way into computer graphics and since then rapidly expanded. The Material Point Method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enable solving of various partial differential equations with ease.

\noindent The Material-Point-Method suffers from high execution times and is thus only viable for hero shots. The method is however highly parallelizable. Thus this thesis proposes how to accelerate the Material-Point-Method using GPGPU techniques.

\noindent Core of the Material-Point-Method are grid and particles transfers that interpolate between the two structures. These transfers are executed multiple times per physical time step. Preprocessing steps might be taken if their computing time is outweighed.

\noindent Deep sorting with counting sort increases coalescing and L2 hit rates. Binning allows to divide the grid into blocks for shared memory filtering techniques. All operations do not rely on fixed bin size. As another preprocessing step only grid blocks are executed which have particles in them.
\vfill
\selectlanguage{german}
\begin{center}
\subsubsection*{Zusammenfassung}
\end{center}
Die Material-Point-Method hat sich in der Computergrafik für physikalische basier\-te Simulationen etabliert. Sie benutzt ein hybrides Modell aus Lagrange-Partikeln und Euler-Gitter. Während Partikel als konsistenter Speicher fungieren, erlaubt das Gitter anfallende partielle Differentialgleichungen effizient zu lösen.

\noindent Die Material-Point-Method unterliegt hohen Berechnungszeiten, die die Methode nur für Hero Shots rentabel macht. Sie ist jedoch hoch-parallelisierbar. Diese Arbeit zeigt, wie die Methode mit GPGPU-Techniken beschleunigt werden kann.

@@

\vfill
\clearpage
\tableofcontents

@@

The simulation of the dynamics of materials is still extremely challenging in different applications and fields. Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible. The Material-Point-Method has already reliable tackled many different models, materials, and even interaction between materials.

Pure Eulerian methods like the Finite-Element-Method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability. Pure Langrangian methods form the end other of the spectrum. Prominent are Smoothed Particle Hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature. Particle-in-Cell(PIC) techniques like the Material-Point-Method are hybrid Lagrangian/Eulerian methods. They try to benefit from the nature of both. Advection on particles is trivial. Thus the Material-Point-Method utilizes Lagrangian particles as a consistent storage of dynamic properties. The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid. The grid is unmoving and does not store any information between frames.

@@

Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as Material-Point-Method extensions. In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out. The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM-Method to get an understanding of the prominent operations. Furthermore GPU intrinsics, metrics and optimization opportunities are discussed. Chapter \ref{sec:implementation} tends to the implementation. The evaluation of the implementation follows up in chapter \ref{sec:eval}. As a conclusion chapter \ref{sec:conclusion} gives a summarizing overview and tends to future work.

@@

The first chapters will derive the Material-Point-Method and necessary extensions for an elastic model. An elastic model is an interesting case displaying much of the needed computations for a typical model. The last chapter \ref{sec:gpgpu} tends to the GPGPU side.

@@

In the Material-Point-Method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}. Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:

@@

The key idea of the Material-Point-Method is to use (Lagrangian) particles as a consistent storage of material properties. All stress based forces are computed on a Eulerian grid however. This grid does not store any material properties and is therefore often referred to as a scratch pad.

@@

In contrast much harder are transfers between the two structures that are essential to the Material-Point-Method:

@@

The number of transfers per frame is much higher than in simple PIC- or SPH-code which are mostly interested in pressure contributions($float$). This motivates making the transfers a very highly optimized operation. The Material-Point-Method requires for an elastic material as derived:
\begin{enumerate}
  \item One P2G-Transfer of mass and momentum $m_i, \boldsymbol{v}_i$(combined $float4$).
  \item Per Conjugate Gradient Iteration:
    \begin{enumerate}
      \item One G2P-Transfer for the $3 \times 3$ Matrix $\boldsymbol{A}_p$
      \item One P2G-Transfer of $\delta \boldsymbol{f}_i$ ($float4$)
    \end{enumerate}
  \item Per Newton Iteration: One G2P-Transfer for $\boldsymbol{F}^{n+1}_p$
  \item One G2P-Transfer of the $3 \times 3$ Matrix $\boldsymbol{B}_p^{n+1}$, the position $\boldsymbol{x}_p^{n+1}$ and the velocity $\boldsymbol{v}_p^{n+1}$
\end{enumerate}
In summary, a lot of data is to be transferred between grid and particles which will only increase with more complex models. In contrast, any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks(chapter \ref{sec:active_blocks}) need to be executed only once every frame. Each transfer can profit from it. Therefore, the transfers are the main focus of this implementation.

@@

By construction the Material-Point-Method tries to avoid matrices on the grid due to storage cost. Conversely, the particles have matrices defined over them, e.g. $\boldsymbol{B}_p$,$\boldsymbol{F}_p$. In conclusion one should motivate approaches where grid nodes correspond to shared memory due to the shared memory size limitation.

@@

\begin{table}[htpb]
  \begin{tabular}{ | l | r | r | r | r |}    \hline
    Method                 &  Area=47.6\% & Area=23.8\% &Area=11.9\%\\\hline
    P2G-Sync		   &  3,375$\mu s$& 3,266$\mu s$ & 3,129$\mu s$\\\hline
    batching = 2           & -&2,942$\mu s$ & 2,745$\mu s$\\\hline
    batching = 4           & -&- &2,622$\mu s$\\\hline
 \end{tabular}
 \caption{P2G-Sync-Transfer of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid. They form a rotated (unsorted) cube with one/two/four particles per cell corresponding to the area of the grid that is covered with 47.6\%/23.8\%/11.9\%. Block size is (4,4,4).}
 \label{table:p2g_area}
\end{table}
The P2G-Atomic- and the even more favorable P2G-Sync-Transfer give great speedups over the global atomic methods. Even in a random case (table \ref{table:p2g_random}) they can outdo the global sorted method despite the high branch divergence. They however shine in the uniform case where speedups of 6.5x and 7.8x respectively can be achieved over the global sorted method.
\clearpage
\section{Conclusion \& Future Work}\label{sec:conclusion}
The derivation of the Material-Point-Method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown. Although the MPM-Method consists of many mathematical tasks four operations on the GPU can be identified. Exclusive grid and particle operations are parallelized with a 1:1 correspondence to threads. The optional use of stream compaction reduces applicable particles/nodes where necessary.

@@

However, before this process can be concluded merge and split techniques need to be taken into consideration. They are designed to augment the numerics of the Material-Point-Method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp. As such they will stray less from the optimistic example. How much they will stray from it is however dependent on the dynamics of the problem.
\clearpage
\begin{appendices}
\section{$D_p$ for Cubic Splines}\label{app:dp_proof}
  \begin{code}
\caption{$\boldsymbol{D}_p$ proof}
  \begin{minted}{python}
import numpy as np
from sympy import *

>>>>

\vfill
\begin{center}
\subsubsection*{Zusammenfassung}
\end{center}
Die Material Point Method hat sich in der Computergrafik für physikalische basier\-te Simulationen etabliert. Sie benutzt ein hybrides Modell aus Lagrange-Partikeln und Euler-Gitter. Während Partikel als konsistenter Speicher fungieren, erlaubt das Gitter anfallende partielle Differentialgleichungen effizient zu lösen.

\noindent Die Material Point Method unterliegt hohen Berechnungszeiten, die die Methode nur für Hero Shots rentabel macht. Sie ist jedoch hoch-parallelisierbar. Diese Arbeit zeigt, wie die Methode mit GPGPU-Techniken beschleunigt werden kann.

@@

\vfill
\begin{center}
\subsubsection*{Abstract}
\end{center}
The Material Point Method is allowing for physically based simulations. It has found its way into computer graphics and since then rapidly expanded. The Material Point Method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enable solving of various partial differential equations with ease.

\noindent The Material Point Method suffers from high execution times and is thus only viable for hero shots. The method is however highly parallelizable. Thus this thesis proposes how to accelerate the Material Point Method using GPGPU techniques.

\noindent Core of the Material Point Method are grid and particles transfers that interpolate between the two structures. These transfers are executed multiple times per physical time step. Preprocessing steps might be taken if their computing time is outweighed.

\noindent Deep sorting with counting sort increases coalescing and L2 hit rates. Binning allows to divide the grid into blocks for shared memory filtering techniques. All operations do not rely on fixed bin size. As another preprocessing step only grid blocks are executed which have particles in them.
\vfill
\clearpage
\tableofcontents

@@

The simulation of the dynamics of materials is still extremely challenging in different applications and fields. Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible. The Material Point Method has already reliable tackled many different models, materials, and even interaction between materials.

Pure Eulerian methods like the Finite-Element-Method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability. Pure Langrangian methods form the end other of the spectrum. Prominent are Smoothed Particle Hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature. Particle-in-Cell(PIC) techniques like the Material Point Method are hybrid Lagrangian/Eulerian methods. They try to benefit from the nature of both. Advection on particles is trivial. Thus the Material Point Method utilizes Lagrangian particles as a consistent storage of dynamic properties. The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid. The grid is unmoving and does not store any information between frames.

@@

Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as Material Point Method extensions. In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out. The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM-Method to get an understanding of the prominent operations. Furthermore GPU intrinsics, metrics and optimization opportunities are discussed. Chapter \ref{sec:implementation} tends to the implementation. The evaluation of the implementation follows up in chapter \ref{sec:eval}. As a conclusion chapter \ref{sec:conclusion} gives a summarizing overview and tends to future work.

@@

The first chapters will derive the Material Point Method and necessary extensions for an elastic model. An elastic model is an interesting case displaying much of the needed computations for a typical model. The last chapter \ref{sec:gpgpu} tends to the GPGPU side.

@@

In the Material Point Method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}. Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:

@@

The key idea of the Material Point Method is to use (Lagrangian) particles as a consistent storage of material properties. All stress based forces are computed on a Eulerian grid however. This grid does not store any material properties and is therefore often referred to as a scratch pad.

@@

In contrast much harder are transfers between the two structures that are essential to the Material Point Method:

@@

The number of transfers per frame is much higher than in simple PIC- or SPH-code which are mostly interested in pressure contributions($float$). This motivates making the transfers a very highly optimized operation. The Material Point Method requires for an elastic material as derived:
\begin{enumerate}
  \item One P2G-Transfer of mass and momentum $m_i, \boldsymbol{v}_i$(combined $float4$).
  \item Per Conjugate Gradient Iteration:
    \begin{enumerate}
      \item One G2P-Transfer for the $3 \times 3$ Matrix $\boldsymbol{A}_p$
      \item One P2G-Transfer of $\delta \boldsymbol{f}_i$ ($float4$)
    \end{enumerate}
  \item Per Newton Iteration: One G2P-Transfer for $\boldsymbol{F}^{n+1}_p$
  \item One G2P-Transfer of the $3 \times 3$ Matrix $\boldsymbol{B}_p^{n+1}$, the position $\boldsymbol{x}_p^{n+1}$ and the velocity $\boldsymbol{v}_p^{n+1}$
\end{enumerate}
In summary, a lot of data is to be transferred between grid and particles which will only increase with more complex models. In contrast, any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks(chapter \ref{sec:active_blocks}) need to be executed only once every frame. Each transfer can profit from it. Therefore, the transfers are the main focus of this implementation.

@@

By construction the Material Point Method tries to avoid matrices on the grid due to storage cost. Conversely, the particles have matrices defined over them, e.g. $\boldsymbol{B}_p$,$\boldsymbol{F}_p$. In conclusion one should motivate approaches where grid nodes correspond to shared memory due to the shared memory size limitation.

@@

\begin{table}[htpb]
  \begin{tabular}{ | l | r | r | r | r |}    \hline
    Method                 &  Area=47.6\% & Area=23.8\% &Area=11.9\%\\\hline
    P2G-Sync		   &  3,375$\mu s$& 3,266$\mu s$ & 3,129$\mu s$\\\hline
    batching = 2           & -&2,942$\mu s$ & 2,745$\mu s$\\\hline
    batching = 4           & -&- &2,622$\mu s$\\\hline
 \end{tabular}
 \caption{P2G-Sync-Transfer of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid. They form a rotated (unsorted) cube with one/two/four particles per cell corresponding to the area of the grid that is covered with 47.6\%/23.8\%/11.9\%. Block size is (4,4,4).}
 \label{table:p2g_area}
\end{table}
The P2G-Atomic- and the even more favorable P2G-Sync-Transfer give great speedups over the global atomic methods. Even in a random case (table \ref{table:p2g_random}) they can outdo the global sorted method despite the high branch divergence. They however shine in the uniform case where speedups of 6.5x and 7.8x respectively can be achieved over the global sorted method.
\clearpage
\section{Conclusion \& Future Work}\label{sec:conclusion}
The derivation of the Material Point Method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown. Although the MPM-Method consists of many mathematical tasks four operations on the GPU can be identified. Exclusive grid and particle operations are parallelized with a 1:1 correspondence to threads. The optional use of stream compaction reduces applicable particles/nodes where necessary.

@@

However, before this process can be concluded merge and split techniques need to be taken into consideration. They are designed to augment the numerics of the Material Point Method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp. As such they will stray less from the optimistic example. How much they will stray from it is however dependent on the dynamics of the problem.
\clearpage
\begin{appendices}
\section{$D_p$ for Cubic Splines}\label{app:dp_proof}
  \begin{code}
\caption{$\boldsymbol{D}_p$ proof}
  \begin{minted}{python}
import numpy as np
from sympy import *
